---
title: "Figure 4"
author: "Ana Luisa"
date: "11/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This RMarkdown will reproduce the panels A to E in Figure 4 (footprinting downstream analysis) for HIV microglia. 
It will import CTCF footprinting resuts, global footprint results, and feature importance from the TAD boundary RF.


##### Import packages
```{r, message=F, echo=FALSE}
addTaskCallback(function(...) {set.seed(123);TRUE}) 
## 
library(GenomicRanges)
library(dplyr)
library(ChIPpeakAnno)

# for plots:
library(ggplot2)
library(ComplexHeatmap)
require(circlize)
require(ggrepel)
require(ggvenn)
```

Paths
```{r}
# Global footprinting results on the 86 TF used for model training in the other cells/tissues
tobias.out.path.tad="data/footprinting/TOBIASsnake_footprintingResults_TAD-TF_2021-09-28.txt"
# Feature importance from the RF model
feat.imp.path = "data/RFmodel/2021-09-28RFmodel_5000bp_featImp.csv"
# Footprinting results on the microglia TF 
tobias.out.path.mglia="data/footprinting/TOBIASsnake_footprintingResults_mglia_2021-09-25.txt"
# CTCF footprints bound for the 3 conditions
ctcf.bound.beds = list.files("data/footprinting/", 
                             pattern = '_bound', 
                             full.names = T)
# Import the same object used for the NMF
path_bins="data/NMF_objects/NMF_input_50000bpBIN_2021-05-27.rds"
```

Import the data  
```{r}
all_tfbs_tads = read.delim(tobias.out.path.tad)
feat.importance = read.csv(feat.imp.path)
all_tfbs_mglia = read.delim(tobias.out.path.mglia)
```


#### Violin plots (Figure 4 A-C)
```{r, fig.height=8, fig.width=10}
# Uninfected/Red
top5updown=rbind(all_tfbs_mglia %>% top_n(uninf_red_change,n = 10) ,
                 all_tfbs_mglia %>% top_n(-uninf_red_change,n = 3))
#

ggplot(all_tfbs_mglia, aes(y=-log10(uninf_red_pvalue), 
                     x=uninf_red_change, 
                     label=ifelse(motif_id %in% top5updown$motif_id, name, ""))) + 
  geom_vline(xintercept=0, color = "darkgrey") +
  geom_point(alpha=ifelse(all_tfbs_mglia$uninf_red_pvalue<=0.05, 1, 0.5),
             color=ifelse(all_tfbs_mglia$uninf_red_change>=0.05, "black", 
                          ifelse(all_tfbs_mglia$uninf_red_change<=-0.01, "#AB0000", "#9F9F9F")),
             size = ifelse(all_tfbs_mglia$uninf_red_change>=0, 
                           (all_tfbs_mglia$uninf_bound/(sum(all_tfbs_mglia$uninf_bound)))*200,
                           (all_tfbs_mglia$red_bound/(sum(all_tfbs_mglia$red_bound)))*200)) +
  scale_x_continuous(expand = expansion(mult = 0.2)) +
  ggrepel::geom_text_repel(data = subset(all_tfbs_mglia, uninf_red_change < 0), 
                           nudge_x = -0.007,
                           nudge_y = -1,
                           segment.alpha  = 0.2,
                           min.segment.length = Inf,
                           direction = "y",
                           hjust = 1,
                           size = 5) +
  ggrepel::geom_text_repel(data = subset(all_tfbs_mglia, uninf_red_change >= 0), 
                           size = 5, 
                           nudge_x = 0.1,
                           segment.alpha  = 0.2,
                           force = 0.5) +
  theme_minimal(base_size = 25) +
  xlim(-0.25, 0.25) +
  labs(x = "Differential binding score",
       y = "-log10(p-value)",
       title = "Uninfected/Latent")
```
```{r, fig.height=8, fig.width=10}
# Uninfected/GFP
top5updown=rbind(all_tfbs_mglia %>% top_n(uninf_gfp_change,n = 10) ,
                 all_tfbs_mglia %>% top_n(-uninf_gfp_change,n = 1))
#

ggplot(all_tfbs_mglia, aes(y=-log10(uninf_gfp_pvalue), 
                     x=uninf_gfp_change, 
                     label=ifelse(motif_id %in% top5updown$motif_id, name, ""))) + 
  geom_vline(xintercept=0, color = "darkgrey") +
  geom_point(alpha=ifelse(all_tfbs_mglia$uninf_gfp_pvalue<=0.05, 1, 0.5),
             color=ifelse(all_tfbs_mglia$uninf_gfp_change>=0.05, "black", 
                          ifelse(all_tfbs_mglia$uninf_gfp_change<=-0.05, "#88CA86", "#9F9F9F")),
             size = ifelse(all_tfbs_mglia$uninf_gfp_change>=0, 
                           (all_tfbs_mglia$uninf_bound/sum(all_tfbs_mglia$uninf_bound))*200,
                           (all_tfbs_mglia$gfp_bound/sum(all_tfbs_mglia$gfp_bound))*200)) +
  # scale_x_continuous(expand = expansion(mult = 0.01)) +
  ggrepel::geom_text_repel(data = subset(all_tfbs_mglia, uninf_gfp_change < 0), 
                           nudge_x = -0.007,
                           nudge_y = -1,
                           min.segment.length = Inf,
                           direction = "y",
                           hjust = 1,
                           size = 5) +
  ggrepel::geom_text_repel(data = subset(all_tfbs_mglia, uninf_gfp_change >= 0), 
                           size = 5, 
                           segment.alpha  = 0.2,
                           nudge_x = 0.1,
                           force = 0.5) +
  theme_classic(base_size = 25) +
  theme_minimal(base_size = 25) +
  xlim(-0.4, 0.4) +
  labs(x = "Differential binding score",
       y = "-log10(p-value)",
       title = "Uninfected/Active")
```

```{r, fig.height=8, fig.width=10}
# Red/GFP
top5updown=rbind(all_tfbs_mglia %>% top_n(red_gfp_change,n = 6) ,
                 all_tfbs_mglia %>% top_n(-red_gfp_change,n = 10))
#

ggplot(all_tfbs_mglia, aes(y=-log10(red_gfp_pvalue), 
                     x=red_gfp_change, 
                     label=ifelse(motif_id %in% top5updown$motif_id, name, ""))) + 
  geom_vline(xintercept=0, color = "darkgrey") +
  geom_point(alpha=ifelse(all_tfbs_mglia$red_gfp_pvalue<=0.05, 1, 0.5),
             color=ifelse(all_tfbs_mglia$red_gfp_change>=0.05, "#AB0000", 
                          ifelse(all_tfbs_mglia$red_gfp_change<=-0.05, "#88CA86", "#9F9F9F")),
             size = ifelse(all_tfbs_mglia$red_gfp_change>=0, 
                           (all_tfbs_mglia$red_bound/sum(all_tfbs_mglia$red_bound)*200),
                           (all_tfbs_mglia$gfp_bound/sum(all_tfbs_mglia$gfp_bound)*200))) +
  # scale_x_continuous(expand = expansion(mult = 0.01)) +
  ggrepel::geom_text_repel(data = subset(all_tfbs_mglia, red_gfp_change < 0), 
                           nudge_x = -0.02,
                           nudge_y = 1,
                           #min.segment.length = Inf,
                           direction = "y",
                           hjust = 1.5,
                           segment.alpha  = 0.2,
                           force = 0.1,
                           size = 5) +
  ggrepel::geom_text_repel(data = subset(all_tfbs_mglia, red_gfp_change >= 0), 
                           nudge_x = 0.05,
                           nudge_y = 4,
                           #min.segment.length = Inf,
                           direction = "y",
                           segment.alpha  = 0.2,
                           hjust = 1,
                           #force = 0.6,
                           size = 5) +
  theme_classic(base_size = 25) +
  theme_minimal(base_size = 25) +
  xlim(-0.2, 0.2) +
  labs(x = "Differential binding score",
       y = "-log10(p-value)",
       title = "Latent/Active")
```


#### Venn diagram for CTCF footprinting  
```{r venn, fig.width=7.5, fig.height=7}
# Generate GRanges from the bound bed files
CTCF.gr = lapply(ctcf.bound.beds, function(x) {
  df = read.delim(x, header = F)
  makeGRangesFromDataFrame(df, seqnames.field = "V1",
                           start.field = "V2", end.field = "V3",
                           keep.extra.columns = T)
})

# quickly rename them as conditions
names(CTCF.gr) <- gsub("CTCF_MA0139.1_", "", 
                       gsub("_bound.bed", "", 
                            basename(ctcf.bound.beds)))

# Save the overlaps using the vennCounts
tovenn = makeVennDiagram(CTCF.gr)$vennCounts

dfvenn = tibble(value = tovenn[2:8,4],
                Active = as.logical(tovenn[2:8,1]),
                Latent = as.logical(tovenn[2:8,2]),
                Uninfected = as.logical(tovenn[2:8,3]))

ggplot(dfvenn, aes(A = Active, B = Latent, C = Uninfected, label = value)) +
  ggvenn::geom_venn(fill_color = c("#429C0E", "#ED2727", "#D3D3D3"),
                    stroke_size = 1, set_name_size = 8, text_size = 5, 
                    stroke_color = "#303030", text_color = "#303030") + 
  theme_void() + 
  coord_fixed()
```

#### CTCF dynamics heatmap  

Import the dataframe with the hotpots  
```{r}
completeMatrix=readRDS(path_bins)

names(completeMatrix) <- gsub(".RPKM", "", 
                            gsub("Overlap", "", names(completeMatrix),ignore.case = T))

# select only the bins with integrations
TargetMatrix <- completeMatrix[completeMatrix$IS>=1,]

# Number of bins with IS is which percentage of total number?
round(nrow(completeMatrix[completeMatrix$IS>=1,])/nrow(completeMatrix)*100, digits = 3)


# how many bins is there with at least n integrations
for (n in c(1:max(completeMatrix$IS))) {
  print(paste0("There are ", 
               nrow(completeMatrix[completeMatrix$IS>=n,]),
               " bins with ", n, " or more IS, ",
               round(nrow(completeMatrix[completeMatrix$IS>=n,])/nrow(TargetMatrix)*100, 
                     digits = 3),
               "% of the total number of bins with IS"))
}

# Let's define hotpots as the set of bins with 2 or more IS?
# It represents ~15% of the total number of bins with IS
hotspots <- makeGRangesFromDataFrame(completeMatrix[completeMatrix$IS>=2,], 
                                     keep.extra.columns = T)

```



Generate the data frames to plot the dynamics heatmap
```{r heatmapCTCF, fig.width=4, fig.height=7}
CTCF.df = read.delim("data/footprinting/CTCF_MA0139.1_overview.txt")

# select only the ones bound in at least one condition
CTCF.bounds = CTCF.df[rowSums(cbind(CTCF.df$uninf_bound, CTCF.df$red_bound, CTCF.df$gfp_bound)) != 0,]
# CTCF.bounds = CTCF.bounds[1:500,] # just for dummy heatmap tests

# check the overlaps with our set of hotspots 
CTCF.bounds$hotspotOverlap = countOverlaps(makeGRangesFromDataFrame(CTCF.bounds,
                                                                   seqnames.field = "TFBS_chr", start.field = "TFBS_start",
                                                                   end.field = "TFBS_end", keep.extra.columns = T), hotspots)

onlyNonIS = as.data.frame(subsetByOverlaps(makeGRangesFromDataFrame(CTCF.bounds,
                                          seqnames.field = "TFBS_chr", start.field = "TFBS_start",
                                          end.field = "TFBS_end", keep.extra.columns = T), 
                 makeGRangesFromDataFrame(completeMatrix[completeMatrix$IS==1,], 
                                                        keep.extra.columns = T)))

CTCF.bounds = anti_join(CTCF.bounds, onlyNonIS)

CTCF.bounds$hotspotOverlap = gsub(1, "Hotspot", 
                                  gsub(0, "No IS", CTCF.bounds$hotspotOverlap))


# select only what we will plot in the body of the heatmap and the annotation
# CTCF.dynam.mat = CTCF.bounds %>% 
#   select(uninf_red_log2fc, uninf_gfp_log2fc, red_gfp_log2fc,hotspotOverlap)

# 
CTCF.bounds$id <- paste0(CTCF.bounds$TFBS_chr, ":", 
                         CTCF.bounds$TFBS_start, "-",
                         CTCF.bounds$TFBS_end)

CTCF.dynam.mat = CTCF.bounds %>%
  select(uninf_score, red_score, gfp_score, 
         hotspotOverlap, id, gene_name, relative_location)
colnames(CTCF.dynam.mat) <- gsub("uninf_score", "Uninfected",
                                 gsub("red_score", "Latent",
                                      gsub("gfp_score", "Active",
                                           colnames(CTCF.dynam.mat))))

# CTCF.dynam.mat = CTCF.bounds %>% 
#   select(uninf_bound, red_bound, gfp_bound) %>% as.matrix()

```
  
  
Plot!  
```{r kmeansCTCF, fig.width=4, fig.height=7}
bound_threshold = rowMeans(cbind(CTCF.df %>% filter(uninf_bound == 1) %>% summarise(Uninf=min(uninf_score)),
                                 CTCF.df %>% filter(red_bound == 1) %>% summarise(Red=min(red_score)),
                                 CTCF.df %>% filter(gfp_bound == 1) %>% summarise(GFP=min(gfp_score))))

## Now, make the heatmaps using the scores by condition
col_fun = circlize::colorRamp2(c(0, # unbound
                                 log(bound_threshold+1), # threshold for bound
                                 max(log(CTCF.dynam.mat[,1:3]+1))), # maximum value of bound
                               c("#FFD073", "#471F6A", "#180529"))

## color scale for the flogFC heatmap
# col_fun = circlize::colorRamp2(c(min(CTCF.dynam.mat[,1:3]), # minimum value of logFC
#                                  0, # no change
#                                  max(CTCF.dynam.mat[,1:3])), # maximum value of logFC
#                                c("#7602B4", "white", "#FF7D0B"))


# color annotation column
listOfcolors = list(Hotspot = c("Hotspot" = "#5F2A6B", "No IS" ="#EBEBEB"))


set.seed(1234)

CTCF.by.cluster <- list()
id.by.cluster <- list()
h = list()
i=1
#ht = list()
# for (i in 1:length(c(3:7))) {
  print(paste("K = 4"))
  h[[i]] = Heatmap(log(as.matrix(CTCF.dynam.mat[,1:3]+1)),
  #h[[i]] = Heatmap(as.matrix(CTCF.dynam.mat[,1:3]),
              border = TRUE, show_row_names = F,
              row_km = 4,
              clustering_distance_rows = "spearman",
              row_km_repeats = 10,
              row_gap = unit(2, "mm"), use_raster = T,
              column_gap = unit(3, "mm"),
              name = "CTCF TFBS bound in \nat least one condition",
              heatmap_legend_param = list(at = c(0, # min(CTCF.dynam.mat[,1:3]),
                                                 log(bound_threshold+1),# 0, # unbound
                                                 max(log(CTCF.dynam.mat[,1:3]+1))), # max(CTCF.dynam.mat[,1:3])), 
                                          # maximum value of bound
                                      labels = c("Unbound", "Threshold", "Bound"), #c("Down", "No change", "Up"),
                                      border = "black",
                                      grid_width = unit(0.7, "cm"),
                                      labels_rot = 270),
              col = col_fun)

  CTCF.by.cluster[[i]] = row_order(h[[i]])
  
  hots.percent = lapply(CTCF.by.cluster[[i]], function(x) {
    df = as.data.frame(t(summary(factor(CTCF.dynam.mat[x,4]))))
    percent.value = df$Hotspot/sum(df$Hotspot, 
                                   df$`No IS`)*100
    print(ifelse(length(percent.value) == 0, paste0(0, '%'), 
                 paste0(round(percent.value, digits = 1), '%')))
  })
  id.by.cluster[[i]] <- lapply(CTCF.by.cluster[[i]], function(x) {
    CTCF.dynam.mat[x,]
  })
  names(id.by.cluster)[i] <- paste("K = 4")
  
  h_annotation <- HeatmapAnnotation(df = data.frame(Hotspot = CTCF.dynam.mat[,4]), 
                                    which = "row",
                                    col = listOfcolors, border = TRUE,
                                    name = "Hotspot overlap",
                                    annotation_legend_param = list(title = "Hotspot overlap", 
                                                                   border = T))
  hpercent = rowAnnotation(foo = anno_empty(border = FALSE, 
                                            width = max_text_width(unlist(hots.percent)) + 
                                              unit(1, "mm")))

  names(h)[i] = paste0("k4")
  # print(h[[i]] ) # + hotpercent)

  clust_cols = RColorBrewer::brewer.pal(7, "PuOr")
  draw(h[[i]] + h_annotation + hpercent) 
  for(o in 1:4) {
    decorate_annotation("foo", slice = o, {
        grid.rect(x = 0, width = unit(1, "mm"), gp = gpar(fill = clust_cols[o], col = NA), just = "left")
        grid.text(paste0(" ", hots.percent[[o]]), 
                  x = unit(1, "mm"), gp = gpar(cex = .8) , just = "left")
    })
}
#}
```

#### Feature importance heatmap  
```{r featIMP, fig.height=7, fig.width=4.2}
# Import feat importance
feat.importance = read.csv(feat.imp.path)

merged.imp = left_join(feat.importance, all_tfbs_tads,
                       by = c("Feature" = "output_prefix")) %>%
  select(Feature, name, Overall,
         uninf_red_change, uninf_red_pvalue,
         uninf_gfp_change, uninf_gfp_pvalue,
         red_gfp_change, red_gfp_pvalue)
merged.imp[is.na(merged.imp)] <- 0
merged.imp = merged.imp[merged.imp$Feature != "ATAC",]
merged.imp$Importance <- rank(-merged.imp$Overall)

## Keep calm and switch the negative to positive values on the uninfected comps to switch comparison baseline
merged.imp$uninf_red_change <- merged.imp$uninf_red_change*-1
names(merged.imp) <- gsub("uninf_red_", "red_uninf_", names(merged.imp))

merged.imp$uninf_gfp_change <- merged.imp$uninf_gfp_change*-1
names(merged.imp) <- gsub("uninf_gfp_", "gfp_uninf_", names(merged.imp))


# make color palette
col_fun = circlize::colorRamp2(c(-0.4, 0, 0.4), c("#5F2A6B", "white", "#DE7A09"))
merged.imp = merged.imp[merged.imp$Importance <= 20,]
Heatmap(as.matrix(merged.imp[,c(4,6,8)]), 
        name = "TFBS change",
        column_labels = c("Latent/Uninf", "Active/Uninf", "Latent/Active"),
        col = col_fun,
        cell_fun = function(j, i, x, y, width, height, fill) {
          pvals = merged.imp[,c(5,7,9)]
          if(pvals[i,j] >= 0.01){
            grid.text(sprintf("n.s", pvals[i,j]), 
                      x, y, gp = gpar(fontsize = 10))
            # } if(merged.imp[i, c(5,7,9)] <= 0.01){
            # grid.text(sprintf("**", merged.imp[i, c(5,7,9)]), 
            #           x, y, gp = gpar(fontsize = 10))
            # } if(merged.imp[i, c(5,7,9)] <= 0.05){
            #   grid.text(sprintf("*", merged.imp[i, c(5,7,9)]), 
            #             x, y, gp = gpar(fontsize = 10))
            # } else{ #(merged.imp[i, c(5,7,9)] > 0.05){
            # grid.text(sprintf("n.s.", merged.imp[i, c(5,7,9)]), 
            #           x, y, gp = gpar(fontsize = 10))
            # }
          }},
        rect_gp = gpar(col = "white", lwd = 2),
        row_labels = merged.imp[,2], #row_names_gp = gpar(hjust=-1), 
        #border = T,
        row_order = rev(order(merged.imp[,3])),
        row_names_side = "left") + 
  rowAnnotation("Predicted \nimportance on \nTAD structure" = anno_barplot(merged.imp$Overall, width = unit(3, "cm"),
                                              bar_width = 1, gp = gpar(col = "white", fill = "#646464"), 
                                              add_numbers = TRUE, border = F), 
                    annotation_name_gp = list(cex = 0.8))
```









```{r}
sessionInfo()
```
