---
title: "Figure 3 (A and B panels)"
author: "Ana Luisa"
date: "10/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This RMarkdown will reproduce the panels A and B in Figure 3 (NMF run and downstream analysis) for HIV microglia analysis. 


##### Import packages
```{r, message=F, echo=FALSE}
addTaskCallback(function(...) {set.seed(123);TRUE}) # keep set.seed = 123 in all functions

library(GenomicRanges)
library(viridis)
library(ComplexHeatmap)
# library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(knitr)
library(tidyverse, lib.loc = "/opt/R/4.0.2/lib/R/library")
library(BiocStyle)
# for NMF
library(ButchR)
library(reticulate)
reticulate::use_condaenv(condaenv = "runningButchR", 
                         conda = "/Volumes/ag-cherrmann/lcosta/miniconda/conda_curry/conda/bin/conda", required = TRUE)
## this environment is supplied in the folder data/envs
```

Paths   
```{r}
# IS
ISpath="data/complete_WT-IScombined_hg38_annotated2021-05-24.rds"
# Paths for the NMF input and output data
NMF_input="data/NMF_objects/NMF_input_50000bpBIN_2021-05-27.rds"
NMF_out="data/NMF_objects/28-06-2021NMF_microglia50000.rds"
```
  
Import the NMF input  
```{r}
completeMatrix=readRDS(NMF_input)

names(completeMatrix) <- gsub(".RPKM", "", 
                            gsub("Overlap", "", names(completeMatrix),ignore.case = T))
## Let's maybe make some identifiable rownames, shall we?
rownames(completeMatrix) <- paste0(completeMatrix$chr, ":",
                                   completeMatrix$start, "-",
                                   completeMatrix$end)
```

Prepare to run the NMF
```{r}
butch.input <- completeMatrix[,-c(1:3,5,13:ncol(completeMatrix))]

dim(butch.input)

k_min <- 2
k_max <- 7
```

Running ButchR  
```{r, warning=F, message=F}
# np_nmf_exp <- run_NMF_tensor(X = limma::normalizeQuantiles(butch.input), ranks = k_min:k_max, method = "NMF", 
#                              n_initializations = 20, extract_features = TRUE)
# 
# OptK(np_nmf_exp) # check the optimal K
# 
# saveRDS(np_nmf_exp,"data/NMF_objects/28-06-2021NMF_microglia50000.rds")

np_nmf_exp <- readRDS(NMF_out)
```  

```{r}
# make the features in the W matrix comparable
np_nmf_exp_norm <- normalizeW(np_nmf_exp)
```

Produces a riverplot to assess the stability of signatures through different k 
```{r riverPlot, message=F, warning=F, error=F, echo=F, results='hide'}
# Make riverplot
plt <- generateRiverplot(np_nmf_exp_norm, edges.cutoff = 0.2)
plot(plt, 
     plot_area = 1, yscale = 0.6, nodewidth = 0.5)
```

```{r}
# Check the resulting H matrix
# H for the optimal decomposition (i.e. the one with the minimal residual) 
# for a specific factorization rank k. 
# The number of rows of the matrix H corresponds to the chosen factorization rank.

np_Hk2 <- HMatrix(np_nmf_exp_norm, k = 6)
class(np_Hk2)

dim(np_Hk2)

# Check the resulting W matrix
# The number of columns of the matrix W corresponds to the chosen factorization rank.
np_Wk2 <- WMatrix(np_nmf_exp_norm, k = 6)
dim(np_Wk2)

# check the W matrix structure
kable(as.data.frame(np_Wk2[1:5, ]))
```



Heatmap visualization for each K
```{r HMatrix, fig.height=5, fig.width=10}
annot_df <- data.frame(Origin = gsub("Mock",
                                     "", colnames(butch.input)))
                       

  nb.cols <- 100
  PuBucontinuous <- colorRampPalette(RColorBrewer::brewer.pal(100, "BuPu"))(nb.cols)

colors.annot = HeatmapAnnotation(df = annot_df, 
                                 col = list(Origin = c("H3K9me2" = "#542788", 
                                                       "H3K27me3" = "#8073ac",
                                                       "H3K9me3" = "#b2abd2",
                                                       "H3K4me1" = "#fee0b6",
                                                       "H3K36me3" = "#d8daeb",
                                                       "H3K27ac" = "#e08214",
                                                       "RNAseq" = "#fdb863",
                                                       "ATAC" = "#b35806")))

#for(ki in k_min:k_max) {
  cat("\n")
  #cat("  \n#### H matrix for k=",  ki, "   \n  ")
  cat("  \n#### H matrix for k=4")
  #plot H matrix
  tmp_hmatrix <- HMatrix(np_nmf_exp_norm, k = 4)
  h_heatmap <- Heatmap(tmp_hmatrix,
                       #col = viridis(100),
                      col = PuBucontinuous,
                       name = "Exposure",
                       clustering_distance_columns = 'pearson',
                       show_column_dend = TRUE,
                       show_column_names = F,
                       top_annotation = colors.annot,
                       show_row_names = FALSE,
                       cluster_rows = FALSE)
  print(h_heatmap)
#}
```
  
```{r}
## Row K-means to determine signature specific features
sig_features <- SignatureSpecificFeatures(np_nmf_exp_norm, 
                                          return_all_features = TRUE)
sig_features <- lapply(sig_features, function(x){
  colnames(x) <- paste0("Sign.", 1:ncol(x))
  x
})

lapply(sig_features, function(x) kable(head(x)))
```


```{r}
# Check the features further
np_specific <- lapply(sig_features, function(x) rownames(x)[rowSums(x) == 1])
np_Wspecific <- lapply(names(np_specific), function(x) {
  i = as.numeric(gsub("k", "", x))
  wmat = WMatrix(np_nmf_exp_norm, k = i)[np_specific[[x]], ]
  colnames(wmat) <- paste0("Sign.", 1:i)
  wmat
})

# normalize exposure score in W matrix across rows
np_Wspecific <- lapply(np_Wspecific, function(x) x/matrixStats::rowMaxs(x))
```
  
Create the global annotation guide for the heatmap
```{r}
heat_anno_rows <- data.frame(bin = rownames(completeMatrix),
                             SE = completeMatrix$SE,
                             IS = completeMatrix$IS)
```  
  
Check the signature features:
```{r WMatrix, fig.height=9, fig.width=5}
# Display selected features on W matrix
library(circlize)
listOfcolors = list(SE = c("no" = "#FFFFFF", "yes" ="#3D3D3D"), 
                    IS = c("no" = "#FFFFFF", "yes" ="#7D3C98"))

# just plot k=4 (2)
vls_by_cluster <- list()
w_heatmap <- list()
for (i in c(2)) {
  locations = data.frame(SE = heat_anno_rows[which(heat_anno_rows$bin %in% 
                                                       rownames(np_Wspecific[[i]])),]$SE,
                         IS = heat_anno_rows[which(heat_anno_rows$bin %in% 
                                                       rownames(np_Wspecific[[i]])),]$IS)
  # change levels
  locations[which(locations$SE >= 1 ),"SE"] <- "yes"
  locations[which(locations$SE ==0 ),"SE"] <- "no"
  
  locations[which(locations$IS >= 1 ),"IS"] <- "yes"
  locations[which(locations$IS ==0 ),"IS"] <- "no"
  
  set.seed(4321) # make sure the cluster order is preserved
  row.annot = HeatmapAnnotation(df = locations, which = "row",
                                col = listOfcolors, border = TRUE, 
                                annotation_legend_param = list(border = T))
  w_heatmap[[i]] <- Heatmap(np_Wspecific[[i]], 
                            col = PuBucontinuous,
                            name = "W matrix",
                            row_km = 5,
                            #clustering_distance_columns = 'pearson',
                            show_column_dend = TRUE,
                            show_column_names = TRUE,
                            show_row_names = FALSE,
                            cluster_rows = TRUE,
                            cluster_columns = FALSE,
                            use_raster = T, 
                            #row_split = ncol(np_Wspecific[[i]]), 
                            row_title = NULL, border = TRUE, 
                            heatmap_legend_param = list(border = T))
  htt = draw(w_heatmap[[i]])
  vls_by_cluster[[i]] = row_order(htt)
  percent.anno = lapply(vls_by_cluster[[i]], function(x) {
    se_val = as.data.frame(t(summary(factor(locations[x,1]))))
    is_val = as.data.frame(t(summary(factor(locations[x,2]))))
    percent.value.se = se_val$yes/sum(se_val$yes,
                                      se_val$no)*100
    percent.value.is = is_val$yes/sum(is_val$yes,
                                      is_val$no)*100
    print(paste0("SE overlap: \n ", round(percent.value.se,
                       digits = 1), '% \n',
                 " IS overlap: \n ", round(percent.value.is,
                       digits = 1), '%'))
  })

  cpercent = rowAnnotation(foo = anno_empty(border = FALSE,
                                            width = max_text_width(unlist(percent.anno)) +
                                              unit(1, "mm")))
  set.seed(4321) # make sure the cluster order is preserved
  draw(w_heatmap[[i]] + row.annot  + cpercent)
  clust_cols = RColorBrewer::brewer.pal(6, "PuOr")
  for(o in 1:5) {
    decorate_annotation("foo", slice = o, {
        grid.rect(x = 0, width = unit(1, "mm"), gp = gpar(fill = clust_cols[o], col = NA), just = "left")
        grid.text(paste0(" ", percent.anno[[o]]),
                  x = unit(.5, "mm"), gp = gpar(cex = .8) , just = "left")
    })}
}
```  
  
```{r}

```



```{r}
sessionInfo()
```

