---
title: "Figure 2 (panel D)"
author: "Ana Luisa"
date: "10/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This RMarkdown will reproduce the panel D in Supplementary Figure 2 (IS vs RMC on genic/intergenic) for HIV microglia analysis. 
Production of the profile plots is a bit time and memory consuming, so prepare yourself. 

##### Import packages
```{r, message=F}
library(GenomicRanges)
library(dplyr)
## for bigwigs:
library(soGGi)
## for plots:
library(ggplot2)
## for genes
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

##### Import data  
Paths
```{r}
hauntedAnalysis=T # if true, will use the phantom sets 

getwd()
# IS
ISpath="data/complete_WT-IScombined_hg38_annotated2021-05-24.rds"

# You will need the merged bigwigs for all the histone marks and atac-seq
H3K27ac.peaks.location1="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_lusic/pipeline_run_PE/Mglia/mock/H3K27ac/bw/Mglia_mock_H3K27ac_RPKM_None_subtract.bw"
H3K27ac.peaks.location2="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_lusic/pipeline_run_SE_H3K27ac/mglia/mock/H3K27ac/bw/mglia_mock_H3K27ac_RPKM_None_subtract.bw"
H3K9me2.peaks.location="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_lusic/pipeline_run_PE/Mglia/mock/H3K9me2/bw/Mglia_mock_H3K9me2_RPKM_None_subtract.bw"
H3K9me3.peaks.location="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_17-02-21/mglia/uninf/H3K9me3/bw/mglia_uninf_H3K9me3_RPKM_None_subtract.bw"
H3K27me3.peaks.location="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_12-10-20/mglia/uninf/H3K27me3/bw/mglia_uninf_H3K27me3_RPKM_None_subtract.bw"
H3K36me3.peaks.location="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_12-10-20/mglia/uninf/H3K36me3/bw/mglia_uninf_H3K36me3_RPKM_None_subtract.bw"
H3K4me1.peaks.location="/Volumes/ag-cherrmann/projects/06_HIV_Microglia/data/chip/microglia_17-02-21/mglia/uninf/H3K4me1/bw/mglia_uninf_H3K4me1_RPKM_None_subtract.bw"

# Import the RMC
if (hauntedAnalysis==T) {
  phantoms=load('data/phantomIS_sets_matchTSSdist_hg38.RData')
}
```
  
Import the IS set from microglia 
```{r}
## IS ranges
IS_ranges <- GRanges(readRDS(ISpath))
how_long = 10000 # how long should the interval of the vicinity of the IS be
```

Use the soGGi function region plot to create the base object of big wig over selected ranges (in this case, IS +/- x)
```{r, message=F}
#genes = ChIPpeakAnno::annoGR(EnsDb.Hsapiens.v86, feature = "gene") ## the original but not working
genes = ChIPpeakAnno::annoGR(TxDb.Hsapiens.UCSC.hg38.knownGene)
genes = genes[seqnames(genes) %in% c(paste0("chr", 1:22), "chrX", "chrY")]

# divide and conquer the RMC
if (hauntedAnalysis==T) {
  Mglia_sets <- list(Microglia = IS_ranges, 
                     PM = phantom_set_mglia[[1]])
  Mglia_sets$PM$Location <- "Intergenic" 
  Mglia_sets$PM[Mglia_sets$PM %in% subsetByOverlaps(Mglia_sets$PM, genes)]$Location <- "Genic" 
  Mglia_sets$PM[Mglia_sets$PM %in% subsetByOverlaps(Mglia_sets$PM, 
                                                    promoters(genes, upstream = 2000, downstream=0))]$Location <- "Genic" 
  
} else {
  Mglia_sets <- list(IS = IS_ranges)
}

```


```{r}
# Active
h3K27ac1 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K27ac.peaks.location1, x, format="bigwig", distanceAround = how_long))
h3K27ac2 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K27ac.peaks.location2, x, format="bigwig", distanceAround = how_long))

h3K36me3 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K36me3.peaks.location, x, format="bigwig", distanceAround = how_long)) 
h3K4me1 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K4me1.peaks.location, x, format="bigwig", distanceAround = how_long)) 

# Repressive
h3K9me3 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K9me3.peaks.location, x, format="bigwig", distanceAround = how_long)) 
h3K27me3 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K27me3.peaks.location, x, format="bigwig", distanceAround = how_long)) 
h3K9me2 <- lapply(Mglia_sets, function(x) 
  regionPlot(H3K9me2.peaks.location, x, format="bigwig", distanceAround = how_long)) 
```


```{r}
plotRegion_remastered_forGroups_genicDivide <- function(ChIPprofileObjectList, range, convert0){
  temp_d <- lapply(ChIPprofileObjectList, function(x) 
    reshape2::melt(data.frame(x@assays@data@listData[[1]],
                              locations = x@rowRanges@elementMetadata@listData[["Location"]])))
  names(temp_d) <- names(ChIPprofileObjectList)
  
  if (convert0 == TRUE) {
    temp_d <- lapply(temp_d, function(x) {
      tt = x
      tt$value[tt$value < 0] <- 0 
      tt })
  }
  
  summaries_list <- lapply(temp_d, function(x){
    n = length(unique(x$variable))
    summaries <- x %>% 
      group_by(variable, locations, .add = T) %>% 
      summarise(mean = mean(value, 
                            na.rm=TRUE),
                sd = sd(value, na.rm=TRUE),
                lower_CI = mean(value, 
                                na.rm=TRUE) - qt(.975,9)*sd(value, 
                                                            na.rm=TRUE)/sqrt(n),
                upper_CI = mean(value, 
                                na.rm=TRUE) + qt(.975,9)*sd(value, 
                                                            na.rm=TRUE)/sqrt(n))
    summaries
  })
  
  binded_summary <- bind_rows(summaries_list, 
                              .id = "Group") 
  
  binded_summary$Set <- paste0(binded_summary$Group, ",",
                               binded_summary$locations)
  
  classygrey = "black"
  pal=c("#1A5276","#512E5F","#D4E6F1", "#D0CCDD")
   ggplot(binded_summary, aes(x = variable, y = mean, 
                              group = Set)) + 
    geom_ribbon(aes(ymin = lower_CI, ymax = upper_CI, 
                    fill=Set), alpha = 0.2) +
    scale_fill_manual(values = pal) +
    scale_color_manual(values = pal) +
    expand_limits(y = 0) +
    geom_line(aes(group = Set, color = Set)) +
    scale_x_discrete(breaks = c("Point_Centre", 
                                paste0("Point_Centre-", range), 
                                paste0("Point_Centre", range)),
                     labels=c("IS", paste0("-", range), 
                              paste0("+", range))) +
    labs(x = "", y = "RPKM") +
    theme_classic(base_size = 20) +
    geom_vline(xintercept = "Point_Centre", 
               linetype="dotted", 
               color = classygrey) +
    theme(plot.margin=unit(c(0.2,1,0.2,0.5),"cm"), 
          axis.title = element_text(colour = classygrey), 
          axis.line = element_line(colour = classygrey), 
          axis.text = element_text(colour = classygrey), 
          legend.position = "right",
          panel.border = element_rect(colour = "black", 
                                      fill=NA, size=2))
}

```
  
  
```{r ChIP, fig.width=10, fig.height=5, message=F}
plotRegion_remastered_forGroups_genicDivide(h3K9me2, how_long, convert0 = TRUE) + 
    ggtitle(paste0("H3K9me2 on the +/-" , how_long, "bp around the IS"))
  
plotRegion_remastered_forGroups_genicDivide(h3K9me3, how_long, convert0 = TRUE) + 
    ggtitle(paste0("H3K9me3 on the +/-" , how_long, "bp around the IS"))
  
plotRegion_remastered_forGroups_genicDivide(h3K27me3, how_long, convert0 = TRUE) + 
    ggtitle(paste0("H3K27me3 on the +/-" , how_long, "bp around the IS"))
  
  ### ----------------
  
plotRegion_remastered_forGroups_genicDivide(list(Microglia=rbind(h3K27ac1$Microglia, h3K27ac2$Microglia),
                                                 PM=rbind(h3K27ac1$PM, h3K27ac2$PM)), how_long, convert0 = TRUE) + 
    ggtitle(paste("H3K27ac on the +/-" , how_long, "bp around the IS"))
  # plotRegion_remastered(h3K27ac1, how_long, convert0 = TRUE) + ggtitle(paste("H3K27ac on the +/-" , how_long, "bp around the IS"))
  # plotRegion_remastered(h3K27ac2, how_long, convert0 = TRUE) + ggtitle(paste("H3K27ac on the +/-" , how_long, "bp around the IS"))
  
plotRegion_remastered_forGroups_genicDivide(h3K4me1, how_long, convert0 = TRUE) + 
    ggtitle(paste0("H3K4me1 on the +/-" , how_long, "bp around the IS"))
  
plotRegion_remastered_forGroups_genicDivide(h3K36me3, how_long, convert0 = TRUE) + 
    ggtitle(paste0("H3K36me3 on the +/-" , how_long, "bp around the IS"))
```


```{r}
sessionInfo()
```




